---
title: Simple Reactable Timeline in R
description: 'Extend reactable framework to accommodate simple Gantt chart'
author: Scott McKenzie
date: '2020-11-22'
slug: reactable-timeline
categories: []
tags: 
  - reactable
  - R
  - timeline
---

```{css echo = FALSE}
.reactable {
  position: relative;
  left: -120px;
}
```


```{r libraries, echo = FALSE, include = FALSE}
library(tidyverse)
library(lubridate)
library(reactable)
library(htmltools)
```


```{r include = FALSE, echo = FALSE}
store_records <- read_csv("https://raw.githubusercontent.com/sccmckenzie/reactable-timeline/master/customer_df.csv")

convert_timestamps <- function(t1, t2) {
  # extract timestamp range
  t01 <- min(t1)
  t02 <- max(t2)
  
  # normalize event timestamps within total range
  left <- time_length(t1 - t01)/time_length(t02 - t01)
  width <- time_length(t2 - t1)/time_length(t02 - t01)
  
  # splice values into list
  out <- list()
  
  for (i in 1:length(left)) {
    out[[i]] <- list(left[i], width[i])
  }
  
  out
}

create_timeline_bar <- function(left = 0, width = "100%", fill = "#00bfc4") {
  left <- scales::percent(left)
  width <- scales::percent(width)
  
  bar <- div(style = list(
    position = "absolute",
    left = left,
    background = fill,
    width = width,
    height = "140%")
  )
  
  chart <- div(style = list(
    flexGrow = 1,
    position = "relative",
    display = "flex",
    alignItems = "center",
    height = "100%"
  ), 
  bar)
  
  div(style = list(
    height = "100%"
  ), 
  chart)
}
```

In this post, I will demonstrate how to create the below timeline using [reactable](https://glin.github.io/reactable/).

```{r echo = FALSE, fig.align = "center"}
df <- store_records %>% 
  mutate(timestamps = convert_timestamps(enter, exit))

reactable(df,
          fullWidth = FALSE,
          width = 905,
          compact = TRUE,
          bordered = TRUE,
          columns = list(
            customer = colDef(width = 100),
            revenue = colDef(width = 100, format = colFormat(digits = 1)),
            enter = colDef(width = 200, format = colFormat(datetime = TRUE)),
            exit = colDef(width = 200, format = colFormat(datetime = TRUE)),
            timestamps = colDef(
              width = 300,
              cell = function(value) {
                create_timeline_bar(left = value[[1]], width = value[[2]])
              }
            )
          )
)
```

```{r echo = FALSE}
br()
```

### Motivation

I've wanted to create an interactive timeline for awhile. There are a lot of great R timeline packages out there, but none of them really fit my needs. 

I'm a huge fan of the reactable package, mainly for its sleek look and customizability. As a result, I had already invested a lot of time integrating "reactables" into my reports. Rather than annihilating my work hitherto, I needed a solution to augment my existing reactables with a graphical timeline.

**The beauty of this solution is that it preserves the tabular framework of reactable.**

First, let's load the data.

```{r message = FALSE}
library(tidyverse)

store_records <- read_csv("https://raw.githubusercontent.com/sccmckenzie/reactable-timeline/master/customer_df.csv")
```

This is a simple dataset that I created for demonstration. It contains store records of customer revenue for 2020-02-03.

* customer: Unique customer ID
* revenue: Total revenue from items purchased by customer
* enter: Time when customer enters store
* exit: Time when customer exits store

```{r}
store_records
```

Our goal is to project `enter` and `exit` into a "timeline column".

### Step 1: Define time transform function

`convert_timestamps` will take `enter` and `exit`, normalize, merge, then return as `list`.

```{r}
library(lubridate) # needed for time_length()

convert_timestamps <- function(t1, t2) {
  # extract timestamp range
  t01 <- min(t1)
  t02 <- max(t2)
  
  # normalize event timestamps within total range
  left <- time_length(t1 - t01)/time_length(t02 - t01)
  width <- time_length(t2 - t1)/time_length(t02 - t01)
  
  # splice values into list
  out <- list()
  
  for (i in 1:length(left)) {
    out[[i]] <- list(left[i], width[i])
  }
  
  out
}
```

Example:
```{r}
store_records %>% 
  mutate(timestamps = convert_timestamps(enter, exit))
```

### Step 2: Define html helper

`reactable()` will feed output from `convert_timestamps` into `create_timeline_bar`, producing our final output.

```{r}
library(htmltools) # needed for div()

create_timeline_bar <- function(left = 0, width = "100%", fill = "#00bfc4") {
  left <- scales::percent(left)
  width <- scales::percent(width)
  
  bar <- div(style = list(
    position = "absolute",
    left = left,
    background = fill,
    width = width,
    height = "140%")
  )
  
  chart <- div(style = list(
    flexGrow = 1,
    position = "relative",
    display = "flex",
    alignItems = "center",
    height = "100%"
  ), 
  bar)
  
  div(style = list(
    height = "100%"
  ), 
  chart)
}
```

### Final product

```{r}
library(reactable)

store_records %>% 
  mutate(timestamps = convert_timestamps(enter, exit)) %>% 
  # ^ created in Step 1
  reactable(fullWidth = FALSE,
            width = 905,
            compact = TRUE,
            bordered = TRUE,
            columns = list(
              customer = colDef(width = 100),
              revenue = colDef(width = 100, format = colFormat(digits = 1)),
              enter = colDef(width = 200, format = colFormat(datetime = TRUE)),
              exit = colDef(width = 200, format = colFormat(datetime = TRUE)),
              timestamps = colDef(
                width = 300,
                cell = function(value) {
                  create_timeline_bar(left = value[[1]], width = value[[2]])
                  # ^ created in Step 2
                }
              )
            )
)
```

