---
title: Crafting an Artificial Dataset
author: Scott McKenzie
date: '2021-02-25'
slug: mopac-dataset
categories: []
tags:
  - R
---

```{r echo = FALSE}
library(ggplot2)
theme_set(theme_minimal())
```


In this post I will explain how I simulated vehicle records for the [Mopac Express Lane](https://www.mobilityauthority.com/traveler-info/open-roads/MoPac-Express) in Austin, Texas.

<!-- Photo goes here -->

# Motivation 

I am currently working towards the release of my second R package, [sift](https://github.com/sccmckenzie/sift). This package is essentially an amalgam of various ideas I've had over the past few years. Since grouping these ideas into one package was very important to me, I needed a common theme to tie all of the functions together.

The dataset created in this post, **`express`**, serves as a realistic case study where one can effectively utilize the full capability of [sift](https://github.com/sccmckenzie/sift).

# Step 1: Reconnaissance

How frequent are various vehicle types on Mopac? Lots of F-150s? Toyota Camrys?

By no means does the intended application of the dataset require 100% accuracy. However, achieving something somewhat realistic was still important to me. After coming up dry on Google, I decided to walk out my front door and get the answer from the primary source.

Discreetly nestled in a grove overlooking Mopac, I aimed my camera at rush hour traffic and let the shutter snap for 2.5 minutes. After repeating this over the course of 1 week, I had collected over 600 frames.

<!-- Put collage here -->

Someday, these photos will serve as an excellent  a Machine Learning algorithm to automatically classify the photos. It took over 8 hours of data entry to transcribe all `r nrow(mopac::rush_hour)` observations.

**I've named this dataset `rush_hour`.**

```{r message = FALSE}
library(tidyverse)

options(readr.default_locale = locale(tz = "US/Central"))

rush_hour <- read_csv("https://raw.githubusercontent.com/sccmckenzie/mopac/master/inst/extdata/rush_hour.csv")
# also available in mopac::rush_hour

rush_hour %>% 
  sample_n(10)
```

If you live in Texas, the below results shouldn't come as a surprise.

```{r eval = FALSE}
rush_hour %>% 
  unite(make_model, make, model, sep = " ") %>% 
  drop_na(make_model) %>% 
  count(make_model) %>% 
  slice_max(order_by = n, n = 10) %>% 
  ggplot(aes(n, fct_reorder(make_model, n))) +
  geom_col()
```


```{r echo = FALSE, fig.height = 3}
rush_hour %>% 
  unite(make_model, make, model, sep = " ") %>% 
  drop_na(make_model) %>% 
  count(make_model) %>% 
  slice_max(order_by = n, n = 10) %>% 
  ggplot(aes(n, fct_reorder(make_model, n))) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  theme(plot.title.position = "plot",
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(r = 250)) +
  labs(title = "Most frequent make/models on Mopac", x = "frequency", y = NULL)
```

Although **`rush_hour`** satisfies the need for realistic vehicle make/model frequencies, it cannot serve as a basis for inferring time-series traffic density in the express lane since the observations were obtained from the Mopac mainlanes.

To avoid further speculation, I made another trip dedicated solely to acquiring express lane timestamps (collected during afternoon rush hour).

```{r}
express_counts <- read_csv("https://raw.githubusercontent.com/sccmckenzie/mopac/master/inst/extdata/express_counts.csv")
```
# Step 2: Defining Scope

Mopac contains an express lane in both directions, stretching 11 miles from downtown Austin to Parmer Lane. Midway between these terminals, there is a checkpoint near Far West Blvd.

* Our mock dataset, **`express`**, will feature vehicle descriptions + timestamps **as if they are captured at the Far West Blvd checkpoint.**
* We'll obtain peak traffic distribution by bootstrapping **`express_counts`**, then adjusting based on City of Austin data.
* Vehicle make/model/color frequencies will be inferred from **`rush_hour`**.[^1]

# Step 3: Initial Traffic Distribution

```{r message = FALSE}
library(lubridate)

# obtain vehicle spacing from rush_hour
set.seed(10)
t_delta <- express_counts %>%
  mutate(t_delta = time_length(time - lag(time))) %>%
  drop_na(t_delta) %>%
  # sprinkle some jitter into timestamps
  # (observations were recorded with 1 sec resolution - this needs to be finer)
  mutate(t_delta = t_delta + 2 * rbeta(n(), 2, 3)) %>%
  pull(t_delta)

head(t_delta)
```

```{r echo = FALSE, fig.height=3}
t_delta %>% 
  qplot(binwidth = 0.25) +
  theme(plot.title.position = "plot",
        plot.title = element_text(margin = margin(0,0,15,0)),
        text = element_text(size = 15),
        plot.margin = margin(r = 15)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Time length between cars [seconds]", title = "t_delta: frequency")
```

To simulate timestamps for **`express`**, we perform bootstrap sampling from the above distribution.

```{r message = FALSE}
# set timeframe (5am - 8pm)
t1 <- 5 
t2 <- 20
total_seconds <- (t2 - t1) * 3600

set.seed(20)
express <- tibble(direction = c("North", "South")) %>%
  # generate temporal vehicle spacing
  rowwise(direction) %>%
  summarize(vehicle_spacing = sample(t_delta, size = total_seconds, replace = TRUE)) %>%
  # add temporal vehicle spacing together
  # & cut off timestamps later than 8pm
  transmute(time = make_datetime(2020, 5, 20, t1, tz = "US/Central") + cumsum(vehicle_spacing)) %>% 
  filter(time < make_datetime(2020, 5, 20, t2, tz = "US/Central"))
  
```

```{r message = FALSE, fig.height = 3}
express %>%
  group_by(direction,
           t15 = floor_date(time, unit = "15 minutes")) %>% 
  summarize(volume = n()) %>% 
  ggplot(aes(t15, volume)) +
  geom_line(aes(color = direction), size = 1.5) +
  scale_y_continuous(limits = c(0, 200)) +
  labs(title = "Simulated Traffic Volume at Mopac Express Lane (Far West)",
       subtitle = "Summed over 15 min intervals",
       x = NULL, y = NULL)
```

Throughout the entire day, the simulated traffic density stays roughly the same - this is unrealistic. For example, around 6:00am, almost nobody uses the express lane.

# Step 4: Exploring City of Austin API

Although Texas state highway traffic count data is elusive (at least at the granularity needed for this application), the City of Austin (COA) provides ample resources on frontage road intersections.

Below we access Mopac & Steck Ave traffic volume measurements from COA open data portal. This intersection is located ~ 1.5mi north of Mopac & Far West, which unfortunately doesn't have records in the [Camera Traffic Counts](https://data.austintexas.gov/Transportation-and-Mobility/Camera-Traffic-Counts/sh59-i6y9) table.

```{r message = FALSE}
library(jsonlite)

api_url <- 'https://data.austintexas.gov/resource/sh59-i6y9.json?atd_device_id=6409&year=2020&month=5&day=20&heavy_vehicle=false'
steck <- fromJSON(api_url) %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  transmute(read_date = as_datetime(read_date, tz = "US/Central"),
            direction,
            movement,
            volume = as.integer(volume))
```

Each row contains measurements summarized over 15 minutes intervals.
```{r}
steck %>% 
  sample_n(5)
```

```{r, message = FALSE, fig.height = 3}
steck %>% 
  group_by(read_date) %>% 
  summarize(volume = sum(volume)) %>% 
  ggplot(aes(read_date, volume)) +
  geom_line(size = 1.5) +
  scale_x_datetime(date_breaks = "3 hours", date_labels = "%R") +
  labs(title = "Traffic Volume at Mopac & Steck Ave",
       subtitle = "Summed over 15 min intervals",
       caption = "Data: City of Austin",
       x = NULL, y = NULL)
```

The above distribution is much more believable: starting around 6am, traffic increases, levels out, gets worse around lunch, and finally recedes after 6pm.

The above chart includes traffic in all 4 directions. Below, we restrict our focus to north & south as these are the only directions that could correspond to cars entering Mopac.

```{r, message = FALSE, fig.height=3}
steck %>% 
  filter(direction %in% c("NORTHBOUND", "SOUTHBOUND")) %>%
  group_by(direction, read_date) %>% 
  summarize(volume = sum(volume)) %>% 
  ggplot(aes(read_date, volume)) +
  geom_line(aes(color = direction), size = 1.5) +
  scale_x_datetime(date_breaks = "3 hours", date_labels = "%R") +
  theme(plot.title.position = "plot") +
  scale_x_datetime(date_breaks = "3 hours", date_labels = "%R") +
  labs(title = "Traffic Volume at Mopac & Steck Ave",
       subtitle = "Summed over 15 min intervals",
       caption = "Data: City of Austin",
       x = NULL, y = NULL)
```

The large discrepancy between north & south may be due to the fact that it's extremely difficult to get on northbound Mopac from Steck Ave. In other words, people may be opting for an alternate access point if they are travelling north.

For the sake of simplicity, we will project the SOUTHBOUND distribution onto our **`express`** dataset (both directions).

```{r}
steck_normalized <- steck %>% 
  filter(direction == "SOUTHBOUND") %>% 
  with_groups(read_date,
              summarize,
              volume = sum(volume)) %>% 
  transmute(id = row_number(),
            read_date,
            volume = volume/max(volume))

set.seed(25)

# join Steck volume with express timestamps
express <- express %>% 
  mutate(id = findInterval(time, steck_normalized$read_date)) %>% 
  left_join(steck_normalized, by = "id") %>% 
  # treat volume as probability of keeping row in express
  rowwise() %>% 
  mutate(keep = sample(c(FALSE, TRUE), prob = c(1 - volume, volume), size = 1)) %>% 
  ungroup() %>% 
  filter(keep) %>% 
  select(direction, time) %>% 
  arrange_all()
```

```{r message = FALSE, fig.height = 3}
express %>%
  group_by(direction,
           t15 = floor_date(time, unit = "15 minutes")) %>% 
  summarize(volume = n()) %>% 
  ggplot(aes(t15, volume)) +
  geom_line(aes(color = direction), size = 1.5) +
  scale_y_continuous(limits = c(0, 200)) +
  labs(title = "Simulated Traffic Volume at Mopac Express Lane (Far West)",
       subtitle = "Summed over 15 min intervals",
       x = NULL, y = NULL)
```

Voila! Admittedly, this is a somewhat hackish method to achieve the desired result. For the intended purpose of this dataset, we don't need to worry about using more sophisticated statistical techniques.

# Step 5: Timestamp Assignment

Now it's time to link actual vehicles to our timestamps.

We are going to add 4 columns:

* License Plate
* Make
* Model
* Color

We'll save generation of the above permutations for the very end. For now, we will denote unique vehicles by integer **`v_id`**. 

```{r}
express <- express %>% 
  mutate(r_id = row_number(), # will save us headache later
         v_id = NA_integer_)
```


Let's outline possible scenarios. Since we cannot rely on COA open data portal to estimate weighting, I've assigned them myself.

* **A**: Vehicle uses Express Lane once **in both directions** (e.g. commuting to & from work) - 50%.
* **B**: Vehicle uses Express Lane thrice **in any combination of directions** (e.g. rideshare) - 10%.
* **C**: Vehicle uses Express Lane once **in one direction** (e.g. traffic wasn't bad enough to justify using Express Lane in both directions) - 40%.

The key challenge will be assigning *realistic* groups of timestamps to one vehicle. A simple call to `sample()` isn't going to cut it.

```{r echo = FALSE, eval = FALSE}
# Scenario A
# Use both directions once
# 50% of timestamps associated with this scenario

n_a <- nrow(express) * 0.5

min_a <- dminutes(30) # this is the fastest someone could "lap" the Far West Blvd checkpoint

set.seed(30)
for (v_id in seq_len(n_a / 2)) {
  
  while(TRUE) {
    a1 <- express %>% 
      filter(is.na(v_id)) %>% 
      mutate(b = time_length(time - floor_date(time, unit = "days"), unit = "hours"),
             wt = dnorm(b, mean = 7, sd = 1)) %>% 
      sample_n(size = 1, weight = wt)
    
    a2 <- express %>% 
      filter(is.na(v_id),
             direction != a1$direction,
             time > a1$time + min_a)
    
    if(nrow(a2) == 0) {
      # message("retry")
      next
    }
    break
  }
  
  a2 <- a2 %>% 
    mutate(b = time_length(time - a1$time, unit = "hours"),
           wt = dnorm(b, mean = 9, sd = 1)) %>% 
    sample_n(size = 1, weight = wt)
  
  express[express$r_id == a1$r_id,]$v_id <- v_id
  express[express$r_id == a2$r_id,]$v_id <- v_id
  
  if (v_id %% 100 == 0) message(v_id)
}
```

```{r eval = FALSE}
hourday <- function(t) {
  time_length(t - make_datetime(2020, 5, 20, tz = "US/Central"), unit = "hours")
}

n_a <- nrow(express) * 0.5 # this needs refinement

north <-   express %>% 
  filter(direction == "North") %>% 
  transmute(north = hourday(time))

south <- express %>% 
  filter(direction == "South") %>% 
  transmute(south = hourday(time))

set.seed(40)
df <- expand_grid(north, south) %>% 
  mutate(t_l = north - south,
         a = if_else(t_l > 0, south, north),
         # a_id = if_else(t_l > 0, s_id, n_id),
         b = if_else(t_l > 0, north, south),
         # b_id = if_else(t_l > 0, n_id, s_id)
         ) %>%  
  filter(t_l > 0.5) %>% 
  nest(data = -a) %>% 
  sample_n(size = n_a / 2, weight = dnorm(a, mean = 7, sd = 1)) %>% 
  unnest(cols = data) %>% 
  anti_join(., ., by = c("b" = "a")) %>% 
  group_by(b) %>% 
  sample_n(size = 1, weight = dnorm(abs(t_l), mean = 10, sd = 2)) %>% 
  group_by(a) %>% 
  sample_n(size = 1, weight = dnorm(abs(t_l), mean = 10, sd = 2)) %>% 
  ungroup()

df %>% 
  ggplot(aes(t_l)) +
  geom_histogram()

```


# Step %: Showtime

* On a highway, moving vehicles tend to coalesce into clusters[^2]. **`sift::kluster`** automatically identifies these natural groupings.

* One would expect to identify many vehicles with one timestamp in each direction (e.g. commuting to work). **`sift::spy`** provides visibility to this conjectural aspect of our dataset.

* What if we wanted to analyze how traffic density changes with Express lane prices? This would require inexact joining with another time-series dataset toll prices. **`sift::interlace`** performs this task in one step.

* Finally, we can use **`sift::sift`** to access chunks of rows surrounding points of interest.

[^1]: Although public service vehicles like metro buses utilize the Express Lane, we will exclude them here.
[^2]: That this doesn't hold true during rush-hour is a moot point as traffic seldom approaches bumper-to-bumper in the Express Lane